### 2009-06-15 paper.tex
paper updates

% figures in figs_plos:
% 1: parent.pdf
% 2: threeway.pdf
% 3: stree-with-structure.pdf
% 4: stree-mutations-with-structures.pdf
% 5: simulations.pdf

@@

We implemented these algorithms for a three-taxon phylogeny.  Using
our implementation on a simple model of pairwise RNA structural
evolution, we simulated data under the model and tested the dependence
of alignment inference and ancestral structure reconstruction accuracy on phylogenetic branch length.
Our results suggest that
we may be able to accurately reconstruct ancestral structures despite
sequence and structural divergence if the evolutionary process is
well-understood.
We discuss the implications of our results for RNA algorithm design
and phylogenetic reconstruction of RNA evolutionary history.

The programs described are available
as part of the \dart\ software package for sequence analysis,
released under the GNU Public License at
\darturl.

@@

This paper focuses on model construction and inference algorithms
rather than the heuristics which will be necessary to make these
algorithms fast enough for analysis of many biological datasets.
As discussed below, the complexity of general inference algorithms is
prohibitively high for many problems of interest.
However, this complexity can be significantly
reduced by incorporating outside knowledge.  For example, if we know
the consensus structure of several sequences or their individual
structures, then we can constrain our algorithms accordingly.
Similarly, we might consider only ancestral structures which are
compatible with a given multiple sequence alignment, or a relatively small set of candidate alignments (as in the ORTHEUS program \cite{PatenHolmesBirney2008}). 
Such constraints are
commonly used by programs for SCFG-based RNA sequence analysis such as QRNA \cite{RivasEddy2001}, Stemloc \cite{Holmes2005} and CONSAN \cite{DowellEddy2006}.
Alignment and structural constraints can be combined \cite{Holmes2005}.

@@

\subsection*{The dependence of reconstruction accuracy on outgroup
  branch length}

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length,
and to investigate the performance of our reconstruction method (when simulating directly from the model)
compared to simpler reconstruction methods that ignore either structure or phylogeny.
(We here use the term ``outgroup'' loosely to denote the
variable-length branch in our three-taxon study,
where the other two branches are held at unit length.)

We simulated the evolution of RNAs under the TKFST model along
three-taxon phylogenies (with one internal node), where we kept the branch lengths of two
sibling species constant and varied the branch length of the
outgroup.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
10]$ bp and stems of lengths $\in [1, 20]$ bp.
We then used \indiegram\ to reconstruct maximum likelihood reconstructions of the ancestral structure and the four-way multiple alignment (ancestor + three descendants).

The results of our simulations, illustrated in \figref{simulations},
demonstrate that the accuracy of both structural
reconstruction and alignment accuracy depend on the outgroup
branch length.  However, while the difficulty of alignment inference
increases in proportion to the outgroup branch length, becoming
near-impossible for very long outgroup branch lengths, the same need
not be true for structural reconstruction accuracy.  As the branch
length of the outgroup increases, the outgroup's structure becomes increasingly decoupled from
that of its ancestor, allowing the scoring scheme to ``ignore'' the
information which it contains.
In the trivial case where the outgroup branch length is zero (so the ancestor is identical to the outgroup),
we can perfectly reconstruct the ancestral structure (as expected),
but we cannot perfectly reconstruct the three-way multiple alignment
 (since there are still two finite-length branches,
 which on average contribute a greater-than-zero rate of unreconstructable mutations).

These results correspond to a best-case scenario in which we
have a perfectly-parametrized evolutionary model and exactly-known descendant structures.
Recent results
demonstrating that inference of sequence alignment is
robust to mis-parametrized probabilistic models \cite{Lunter2007}
suggest that the same may be true for probabilistic inference of
ancestral structures.
%The effect of imperfect knowledge of the ancestral structures has not been explored.
% [commented out because the following tests DO address this in a roundabout way; what we mean is that indiegram has not been tested when leaf structures are unknown]

To compare \indiegram\ to simpler reconstruction methods, we also measured the alignment accuracy using various simpler statistical reconstruction techniques.
We evaluated three maximum likelihood inference algorithms that use simpler gap-aware phylogenetic models,
and one posterior decoding algorithm that uses a non-phylogenetic sum-of-pairs scoring scheme.
These results are plotted on \figref{simulations} alongside the \indiegram\ accuracies.
In detail, the reconstruction methods used were

\begin{description}
\item{TKF91 ML:} with the TKF91 model \cite{}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
The correct phylogenetic tree and branch lengths were supplied (as they were for the \indiegram\ benchmark).
The insertion, deletion and substitution rates for the TKF91 model were set equal to those of the loop submodel of TKFST.
\item{Long Indel ML:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
The correct phylogenetic tree and branch lengths were supplied.
The deletion and substitution rates were set equal to those of the loop submodel of TKFST.
The mean indel length was set equal to $D_l$, the mean number of bases that are created/removed by an insertion/deletion in the loop submodel of TKFST;
the mean equilibrium sequence length (and thereby the insertion rate) was equal to $B_l$, the mean number of bases in TKFST at equilibrium
(\secname{A simple model of RNA structural evolution} has formulae for these quantities in terms of the TKFST rate parameters).
\item{Stemloc ML:} the \stemloc\ program was used to align the three sequences via single-linkage clustering with a Pair SCFG \cite{Holmes2005}.
The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
\item{Stemloc AMA:} the \stemloc\ program was used to align the three sequences in AMA mode, using posterior decoding to optimize a sum-over-pairs expected alignment accuracy \cite{BradleyPachterHolmes2008}.
The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
\end{description}

We did not measure the accuracy of the reconstructed ancestral structure in these tests (as we did in the \indiegram\ test),
since \handel\ does not consider structures, whereas \stemloc\ and \stemlocama\ do not provide ancestral reconstructions.

{\bf SUMMARY OF THESE RESULTS}
% TKF is less accurate than indiegram, although for short branch lengths it is almost identical; the TKF and indiegram accuracies diverge at longer branch lengths
% Long indel?
% Stemloc?
% StemlocAMA?

@@

@@

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/simulations.pdf}
  \caption{
    \textbf{Dependence of reconstruction accuracy on outgroup branch
      length.}
    We investigated the dependence of structural reconstruction and
    alignment accuracy on outgroup branch length by simulating the
    evolution of three structural RNAs under our model.  The two
    sister species were equidistant from the ancestral sequence, with
    a branch length of 1, corresponding to one expected
    substitution per site in loop sequence.  We allowed the outgroup
    branch length to vary between $[0, 2.5]$.  We divided this interval
    into 25 bins and simulated 25 alignments for each bin.
    Parameters used were $\lambda_l = 0.025$ and $\mu_l = 0.03$ for
    loop sequence and $\lambda_l = 0.007$ and $\mu_l = 0.01$ for stem
    sequence; the probability of a stem insertion was $0.1$.  These
    yielded a mean loop length of 5 bp and a mean stem length of 2.33 bp,
    with $\sim 0.3$ substructure indels per alignment.
    We selected alignments to reconstruct by requiring
    that there be at least two ancestral stems, loops of length $\in [3,
    10]$ bp and stems of length $\in [1, 20]$ bp; to reduce the complexity of
    our algorithms we additionally required that the sequences have
    lengths $\in [30, 70]$ bp.
    ``Structural reconstruction accuracy'' is the fraction of
    base-pairs in the ancestral structure present in the
    maximum-likelihood reconstruction thereof; ``Alignment accuracy''
    is a measure of sequence alignment accuracy which takes both
    aligned and unaligned (indels) nucleotides into account,
    qualitatively corresponding to giving equal weight to sensitivity
    and specificity (see
    \cite{SchwartzMyersPachter2006} for a precise definition).
}
  \figlabel{simulations}
\end{figure}

>>>>

% figures in figs_plos:
% 1: parent.pdf
% 2: threeway.pdf
% 3: stree-with-structure.pdf
% 4: stree-mutations-with-structures.pdf
% 5: perfect.pdf
% 6: ama.pdf

@@

We implemented the above algorithms for a simple model of pairwise RNA structural evolution;
in particular the algorithms for maximum likelihood (ML) alignment of three known RNA structures an a known phylogeny
and inference of the common ancestral structure.
We compared this ML algorithm to a variety of related, but simpler, techniques,
including ML alignment algorithms for simpler models that omitted various aspects of the full model
and also a heuristic (sum-over-pairs) posterior-decoding alignment algorithm for one of the simpler models.

Our results suggest that ML algorithms incorporating basepair structure are the most likely to estimate the alignment perfectly.
However, the heuristic posterior-decoding algorithm scores higher on finer-grained alignment accuracy metrics.
Since the posterior-decoding heuristic is substantially faster than exact phylogenetic ML,
this further motivates the use of sum-over-pairs (and approximate sum-over-pairs) heuristics
for paleogenetics and other alignment problems.

@@

This paper focuses on model construction and inference algorithms
rather than the heuristics which will be necessary to make these
algorithms fast enough for analysis of many biological datasets.
As discussed below, the complexity of general inference algorithms is
prohibitively high for many problems of interest.
However, this complexity can be significantly
reduced by incorporating outside knowledge.  For example, if we know
the consensus structure of several sequences or their individual
structures, then we can constrain our algorithms accordingly.
Similarly, we might consider only ancestral structures which are
compatible with a given multiple sequence alignment, or a relatively small set of candidate alignments (as in the ORTHEUS program \cite{PatenHolmesBirney2008}). 
Such constraints are
commonly used by programs for SCFG-based RNA sequence analysis such as QRNA \cite{RivasEddy2001}, \stemloc\ \cite{Holmes2005} and CONSAN \cite{DowellEddy2006}.
Alignment and structural constraints can be combined \cite{Holmes2005}.

@@

\subsection*{Comparison of alignment methods}

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study of the dependence of reconstruction accuracy on outgroup branch length,
with the further goal of comparing the performance of our reconstruction method (when simulating directly from the model)
to simpler reconstruction methods that ignore either structure or phylogeny.
(We here use the term ``outgroup'' loosely to denote the
variable-length branch in our three-taxon study,
where the other two branches are held at unit length.)

We simulated the evolution of RNAs under the TKFST model along
three-taxon phylogenies (with one internal node), where we kept the branch lengths of two
sibling species constant and varied the branch length of the
outgroup between $[0, 2.5]$ at steps of size $0.1$.
    Parameters used in the simulation were $\lambda_l = 0.025$ and $\mu_l = 0.03$ for
    loop sequence and $\lambda_l = 0.007$ and $\mu_l = 0.01$ for stem
    sequence; the probability of a stem insertion was $0.1$.  These
    yielded a mean loop length of 5 bp and a mean stem length of 2.33 bp,
    with $\sim 0.3$ substructure indels per alignment.
    We selected alignments to reconstruct by requiring
    that there be at least two ancestral stems, loops of length $\in [3,
    10]$ bp and stems of length $\in [1, 20]$ bp; to reduce the complexity of
    our algorithms we additionally required that the sequences have
    lengths $\in [30, 70]$ bp.

We then attempted three-way multiple alignment (and, in some cases, reconstruction of the ancestor)
using a variety of statistical inference algorithms.
We sought insight as to the relative importance of the following factors in reconstructing ancestral RNA:
(i) modeling the secondary structure;
(ii) modeling the phylogenetic topology \& branch lengths;
(iii) using posterior-decoding algorithms to maximize the expected alignment {\em accuracy},
rather than picking the single {\em most likely} alignment \cite{Holmes98,SchwartzPachter2007,BradleyPachterHolmes2008}.

The alignment programs we used in this benchmark were
\indiegram\ (exact ML inference of alignment and ancestral structure, given phylogeny, descendant structures and correct model);
\stemloc\ (a greedy ML heuristic, ignoring phylogeny in favor of a single-linkage clustering of the descendant structures);
\stemlocama\ (a posterior-decoding heuristic, maximizing the alignment's {\em expected accuracy} rather than its {\em likelihood});
and
\handel\ (ML alignment under various indel models that ignore secondary structure completely).
In detail, the reconstruction methods were

\begin{description}
\item{\stemloc:} the \stemloc\ program was used to align the three sequences via single-linkage clustering with a Pair SCFG \cite{Holmes2005}.
The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
{\bf This model improves on the previous model (Long Indel) by introducing RNA secondary structure, though it no longer models phylogeny as rigorously as the Long Indel model} (it uses single-linkage clustering instead of a true phylogeny).
\item{\stemlocama:} the \stemloc\ program was used to align the three sequences in ``sequence annealing'' mode,
 a posterior decoding method that attempts to optimize AMA, a sum-over-pairs alignment accuracy metric \cite{BradleyPachterHolmes2008}.
The structures of the leaf sequences were provided, but not the phylogenetic branch lengths.
{\bf This program uses the previous model but instead of maximizing likelihood, it attempts to maximize an alignment accuracy metric.}
\item{TKF91:} with the TKF91 model \cite{ThorneEtal91}, the \handel\ package \cite{HolmesBruno2001,Holmes2003,Holmes2007} was used to align the three extant sequences and reconstruct the ancestor.
The correct phylogenetic tree and branch lengths were supplied (as they were for the \indiegram\ benchmark).
The insertion, deletion and substitution rates for the TKF91 model were set equal to those of the loop submodel of TKFST.
{\bf This may be understood as a naive reconstruction that completely ignores the stem sub-model of TKFST.}
\item{Long Indel:} with a single-event trajectory approximation to the long indel model \cite{MiklosLunterHolmes2004}, the \handel\ package was used to align the three extant sequences and reconstruct the ancestor.
The correct phylogenetic tree and branch lengths were supplied.
The deletion and substitution rates were set equal to those of the loop submodel of TKFST.
The mean indel length was set equal to $D_l$, the mean number of bases that are created/removed by an insertion/deletion in the loop submodel of TKFST;
the mean equilibrium sequence length (and thereby the insertion rate) was equal to $B_l$, the mean number of bases in TKFST at equilibrium
(\secname{A simple model of RNA structural evolution} has formulae for these quantities in terms of the TKFST rate parameters).
{\bf This model improves on the previous model (TKF91) by introducing affine gap penalties.}
\end{description}

We focused primarily on alignment accuracy, under the simplifying assumption that this will be a good indicator for
(and perhaps the primary determinant of) the accuracy of ancestral reconstruction.

We first consider the {\em perfect alignment rate}; that is, the number of times each method gets the alignment exactly correct.
Theory predicts that ML inference, using the correct model and parameters, should maximize the perfect alignment rate.
Inspecting \figref{perfect}, we find this to be the case, except when the outgroup is very distant
and the best perfect alignment rate drops below 0.05 (at which point 125 trials may not be sampling it sufficiently).
We also note that the ML version of \stemloc\ is near-optimal, and slightly better than the posterior-decoding \stemloc-AMA.
Finally, we note that the structure-blind models (TKF91 and Long Indel) perform consistently worse than the structure-aware methods,
with no decisive advantage shown by either TKF91 (linear gap penalty) or Long Indel (affine gap penalty).

A subtly different trend is apparent when we look at the accuracy of the returned alignment,
measured not by the all-or-nothing metric of whether the alignment is {\em perfect},
but rather by a metric that count the number of correct alignment columns.
\figref{ama} shows the {\em Alignment Metric Accuracy} (AMA),
a generalization of the {\em Sum-of Pairs Score} (SPS)
which measures the proportion of residues which are correctly aligned, averaged over all pairs of sequences \cite{SchwartzMyersPachter2006}.
Here, we see that \stemlocama---which attempts to find the alignment with maximum expected AMA---edges out both
\indiegram\ and the ML version of \stemloc.
The decision-theoretic scoring scheme of \stemlocama\ appears to offset the minor incorrectness of its model.

Again, when looking at the AMA statistics, the most important factor is clearly the incorporation of some form of basepair structure
into the model: \handel, whether using TKF91 or the Long Indel model, performs much worse than the SCFG-based methods.
This is to be expected: whenever a basepair-aware method aligns one half of a basepair, it gets the other nucleotide
correctly aligned for free.

@@

Given the excellent performance of \stemlocama's sequence annealing, particularly when measured using its own scoring metric (AMA),
such posterior-decoding methods should also be considered for reconstruction.

@@

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/perfect.pdf}
  \caption{
    \textbf{Dependence of perfect alignment rate on outgroup branch
      length.}
    We investigated the perfect alignment rate
    (the number of times that a given alignment program estimates the alignment 100\% correctly, with no errors)
    as a function of outgroup branch length, simulating the
    evolution of three structural RNAs under our model.  The two
    sister species were equidistant from the ancestral sequence, with
    a branch length of 1, corresponding to one expected
    substitution per site in loop sequence.  We allowed the outgroup
    branch length $t$ to vary between $[0, 2.5]$.  We selected 25 equally spaced values of $t$ in this range, spaced $0.1$ apart,
    and simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
    Since the perfect alignment rate is rather low, we further aggregated the $t$-values for this plot into bins of five;
    thus, for example, the bin named ``$0 to 0.5$'' includes $t \in \{ 0, 0.1, 0.2, 0.3, 0.4 \}$.
    The vertical axis counts the number of perfect alignments in each bin, out of a total of 125 trials per bin.
}
  \figlabel{perfect}
\end{figure}

% figure produced automatically with simulations/plots.r from simulations/inferred/indiegram.analysis
\begin{figure}[!ht]
  \centering
  % \includegraphics[width=5in]{figs_plos/ama.pdf}
  \caption{
    \textbf{Dependence of reconstruction accuracy on outgroup branch
      length.}
    We investigated the dependence of structural reconstruction and
    alignment accuracy on outgroup branch length by simulating the
    evolution of three structural RNAs under our model.  The two
    sister species were equidistant from the ancestral sequence, with
    a branch length of 1, corresponding to one expected
    substitution per site in loop sequence.  We allowed the outgroup
    branch length $t$ to vary between $[0, 2.5]$.  We selected 25 equally spaced values of $t$ in this range, spaced $0.1$ apart,
    and simulated 25 alignments for each value of $t$, using TKFST model parameters described in the text.
    The Alignment Metric Accuracy (AMA) is, roughly, the proportion of residues that are correctly
    aligned, averaged over all pairs of sequences
    (see \cite{SchwartzMyersPachter2006} for a precise definition; we used AMA with Gap Factor 1).
}
  \figlabel{ama}
\end{figure}
